Graded Call-By-Push-Value

### 2020-07-09 popl21/popl21.tex
[ popl21 ] introduction

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
Call-by-push-value (CBPV) is a simply typed lambda calculus that polarizes
types into value and computation types and can thus express both
call-by-name and call-by-value evaluation in the presence of effects.
Semantically, effects are modeled by a monad, and computation types as
algebras over this monad.
Effect type systems usually express more information than the presence
of an effect; often effects are categorized by pre-ordered monoid
where the monoid operation represents accumulation of effects and the
order expresses effect subsumption, in analogy to subtyping.
In this work, a \emph{graded} version of CBPV is presented where the typing
of computations likens effect typing.  Semantically, computation types
are then represented as graded monad algebras.
Further we present a version of CBPV which additionally has graded coeffects.
\end{abstract}

@@

%\newcommand{}{\ensuremath{}}
\newcommand{\bla}{\ensuremath{\mbox{$$}}}
\newcommand{\ie}{\emph{i.e.}}
\newcommand{\eg}{\emph{e.g.}}
\newcommand{\Eg}{\emph{E.g.}}
\newcommand{\loccit}{\emph{loc.\,cit.}}
\newcommand{\cf}{cf.\ }
\newcommand{\den}[2][]{\llbracket#2\rrbracket^{#1}}
\newcommand{\dent}[1]{\llparenthesis#1\rrparenthesis}
\newcommand{\To}{\ensuremath{\Rightarrow}}
%\newcommand{\todot}{\stackrel\cdot\to}
\newcommand{\todot}{\mathbin{\dot{\to}}}
\newcommand{\bN}{\ensuremath{\mathbb{N}}}
\newcommand{\dom}{\mathop{\mathrm{dom}}\nolimits}
\newcommand{\Pot}[1]{\mathcal{P}\,#1}
%\newcommand{\defiff}{:\iff}
\newcommand{\defiff}{\mathrel{{{:}{\Longleftrightarrow}}}}
\newcommand{\subst}[3]{#3[#1/#2]}
\newcommand{\qsubst}[4]{#4[#1/#2#3]}

@@

@@

Call-by-push-value (CBPV) is a simply typed lambda calculus that polarizes
types into value and computation types and can thus express both
call-by-name and call-by-value evaluation in the presence of effects.
Semantically, effects are modeled by a monad, and computation types as
algebras over this monad.
Effect type systems usually express more information than the presence
of an effect; typically effects are structured as pre-ordered monoid
where the monoid operation represents accumulation of effects and the
order subsumption, in analogy to subtyping.
In this work, a \emph{graded} version of is presented where the typing
of computations likens effect typing.  Semantically, computation types
are then represented as graded monad algebras.
Further we present a version of CBPV which additionally has graded coeffects.

\paragraph*{Preliminaries.}  The reader should bring some elementary
knowledge of category theory, such as the interpretation of
simply-typed lambda calculus in cartesian-closed categories.  However,
we try to be gentle with categorical concepts such as monads and
recapitulate their definition where needed.  In many cases, it is
sufficient to think in terms of the category $\SET$ where objects are
sets and morphisms functions between sets, or in the functor category
$[\CC \to \SET]$ where objects are monotone families of sets
$(A_i)_{i:\CC}$ indexed by objects $i$ of $\CC$ such that
$\CC_i \to \CC_j$ in $\SET$ for $i \to j$ in $\CC$ and morphisms are
natural transformations $(f_i : A_i \to B_i)_{i:\CC}$.

\section{An effect-graded version of CBPV}

@@

\newcommand{\tprint}{\mathsf{print}}
\newcommand{\print}[2]{\tprint\,#1.\,#2}
\newcommand{\toutput}{\mathsf{output}}
%\newcommand{\output}[2][]{\toutput_{#1}\,#2}
\newcommand{\tthrow}{\mathsf{throw}}
\newcommand{\throw}[1]{\tthrow\,#1}
\newcommand{\tcatch}{\mathsf{catch}}
\newcommand{\catch}[3]{#1\,\tcatch\,#2 \mapsto #3}
\newcommand{\traise}{\mathsf{raise}}
%\newcommand{\raise}[1]{\traise\,#1} %clash
\newcommand{\thandle}{\mathsf{handle}}
\newcommand{\handle}[4][]{\thandle_{#1}\,#2\,#3\,#4}

@@

A \emph{monoidal} comonad implements an morphism $\tzip$ that combines
a tuples of comonadic values into one comonadic tuple:
\[
\begin{array}{lcl}
  \tzip & : & \tupt {i:I} {(\D\,B_i)} \to \D(\tupt I B) \\
  \tzip\,(s_i)_{i:I}\,n & = & (s_i\,n)_{i:I} \\
\end{array}
\]
Many other comonads are monoidal,
like the store comonad $(S \To \_) \times S$,
though not all, \eg, the context comonad $S{\times}\_$.
In this article, we consider only monoidal comonads.

@@

Because comonad coalgebras are closed under products for monoidal
coalgebras, contexts $\Gamma$ are interpreted as comonad coalgebras
$\den\Gamma$.  In the presence of a comonad interpretation of $\Box$,
the operations $\tthunk$ and $\tforce$ are no longer the identity, but
``generalized cobind'' and $\extract$:
\[
\begin{array}{lllcl}
 \rintro\Box
  & \ru{\oCompTy \Gamma t N
      }{\ValTy \Gamma {\thunk t} {\Box N}}
  & \dent{\thunk t} & = & \fmap[\D]{\dent t} \comp \expose[\den\Gamma]
\\[3ex]
 \relim\Box
  & \ru{\ValTy \Gamma v {\Box N}
      }{\oCompTy \Gamma {\force v} N}
  & \dent{\force v} & = & \textract \comp \dent v
\end{array}
\]
The map $\dent\tthunk : \Hom A B \to \Hom A {\D\,B}$ generalizes
$\tcobind : \Hom{\D\,A} B \to \Hom{\D\,A}{\D\,B}$ to $\D$-coalgebras
$A$ in the same way that the monadic
$\tbind : \Hom{A}{\T\,B} \to \Hom{\T\,A}{\T\,B}$ is generalized to
$\Hom{A}{B} \to \Hom{\T\,A}{B}$ for $\T$-algebras $B$ in CBPV.

@@

Graded comonads have been utilized to give semantics to
context-dependent computation \citep{orchard:icfp14}.  For grading,
\loccit uses a \emph{resource algebra} in form of a preordered
semiring
$(\R,{+},0,{\cdot},1,\leq)$.
% $(\R,\_{+}\_,0,\_{\cdot}\_,1,\leq)$.
%
The semantics is \emph{resource aware} and thus not constructed in a
cartesian-closed category but in a symmetric monoidal closed category
$(\CC,\otimes,\I,\lolli)$.  We overload the operators on
exponentials, $\Lambda : \Hom {C \otimes A} B \to \Hom C {A \lolli B}$
and $\teval : \Hom{(A \lolli B) \otimes A} B$.  The tensor product
$\otimes$ is a bifunctor and it is customary to overload $\otimes$ for
the functorial action
$f_1 \otimes f_2 : \Hom{A_1 \otimes A_2}{B_1 \otimes B_2}$ where
$f_i : \Hom{A_i}{B_i}$.  We shall also use $I$-ary products
$\tupt {i:I} {A_i}$ and their functorial action $\tupt {i:I} {f_i}$.
Further, the symmetric monoidal structure is usually witnessed by
isomorphisms for left $\lambda_A : \Hom {\I \otimes A} A$ and right
unit $\rho_A : \Hom {A \otimes I} A$, associativity
$\Ga_{A,B,C} : \Hom{A \otimes (B \otimes C)} {(A \otimes B) \otimes
  C}$
and swap $\sigma_{A,B} : \Hom {A \otimes B} {B \otimes A}$.  We shall
summarize combinations of these isomorphisms under the name
$\isot : \Hom {\tupt I A} {\tupt J B}$ when the multisets
$\{A_i\}_{i:I}$ and $\{B_j\}_{j:J}$ coincide modulo addition and
deletion of $\I$s.

As a running example for resource accounting,
we use the semiring $\R = \Pot\bN \setminus \emptyset$ of
\emph{multiplicities} ordered by $\supseteq$ with pointwise sum and
product, \eg, $r + s = \{ n + m \mid n \in r, m \in s \}$.
Subsemirings of $\R$ have been used for quantitative typing
\cite{sergeyVytiniotisPeytonJones:popl14,mcBride:wadler60,atkey:lics08}.
The order expresses precision of the quantities, \eg,
$\{1\} \geq \{0,1\} \geq \bN$ states that linear use of a variable is more informative than affine use than unrestricted use.

@@

Objects of $\CC$ are functors $A : (\W,\wleq) \to \SET$ and morphisms
$f : \Hom A B$ are natural transformations
$(f_w : A_w \to B_w)_{w:W}$, \ie,
$B_{w \wleq w'} \comp f_w = f_{w'} \comp A_{w \wleq w'}$.  The tensor
$A \otimes B$ is Day's convolution
$(A \otimes B)_w = \bigcup_{w_1 \oplus w_2 \wleq w} (A_{w_1} \times B_{w_2})$
% $(A \otimes B)_w = \bigcup_{w_1,w_2 \mid w \geq w_1 \oplus w_2} A_{w_1} \times A_{w_2}$
with unit $\I_w = \bigcup_{\emp \wleq w} 1$.  The unit $\I$ is
constructible at world $w$ from nothing (set $1$) if $\emp \wleq w$,
\ie, if the world $w$ includes the choice of using no resources.  A
tensor $A \otimes B$ is constructible at world $w$ if $w$ includes the
choice to split the resources into $w_1$ and $w_2$ to construct $A$
and $B$, resp.  The exponential $A \lolli B$ is determined by currying
$C \otimes A \lolli B \cong C \lolli (A \lolli B)$ and given by
$(A \lolli B)_w = \bigcap_{w_1 \oplus w \wleq w_2} (A_{w_1} \to
B_{w_2})$.

The thus constructed symmetric monoidal closed category $\CC$ has a
$\R$-graded comonad
\[
\begin{array}{lcl}
  \D & : & (\R,\leq) \to [\CC \to \CC] \\
(\D[r] A)_w & = &
  \bigcup_{rw' \wleq w} A_{w'}
  % \bigcup_{w' \mid w \wleq rw'} A_{w'}
\end{array}
\]
implementing trivially the following natural transformations.  Herein, we use the
pointwise tensor product of functors
$(\dot\otimes_{i:I}F_i)\,A = \otimes_{i:I}(F_i\,A)$ in
$[\CC \to \CC]$, in particular $\dot\I\,A = \I$.
\[
\begin{array}{lcl@{~}c@{~}l}
  \textract   & : & \D[1]   & \todot & \Id \\
  \tdisplay & : & \D[rs]  & \todot & \D[r] \comp \D[s] \\
  \tdrop    & : & \D[0]   & \todot & \dot{\I} \\
  \tduplicate  & : & \D[r+s] & \todot & \D[r] \mathbin{\dot\otimes} \D[s] \\
  % \tdrop    & : & \Hom {\D[0]\,A} A \\
  % \tduplicate  & : & \Hom {\D[r+s]\,A} {\D[r]\,\A \otimes \D[s]\,A} \\
\end{array}
\]
For example, $\tduplicate$ takes
$a \in \bigcup_{rw' \oplus sw' \wleq w} A_{w'}$ to
$(a,a) \in \bigcup_{w_1 \oplus w_2 \wleq w} \bigcup_{rw_1' \leq w_1}
\bigcup_{sw_2' \leq w_2} A_{w'_1} \times A_{w'_2}$.
The intermediate worlds are $w_1' = w_2' = w'$ and $w_1 = rw'$ and $w_2 = sw'$.

@@

A $\D$-coalgebra is a functor $A : (\R,\leq) \to \CC$ with family of
morphisms $\expose[r,s] : A_{rs} \to \D[r]\,A_s$ natural in $r$ and
$s$.  For the $\D$-coalgebra $\D[\_]\,B$ the family $\expose$ is just
$\tdisplay$, and the suitably generalizable laws for $\tdisplay$ are
required to hold for $\texpose$:
\[
\xymatrix@C=10ex{
A_r  \ar[d]_{\texpose} \ar@{=}[dr]
\\
\D[1]\,A_r \ar[r]^{\textract}
  & A_r
}
\qquad
\xymatrix@C=12ex{
A_{qrs} \ar[d]_{\texpose} \ar[r]^{\texpose}
  & \D[qr]\,A_s \ar[d]^{\tdisplay}
\\
\D[q]\,A_{rs} \ar[r]^{\tfmap\,\texpose}
  & \D[q]\,(\D[r]\,A_s)
}
\]
Further, $A$ needs to map the additive monoidal structure $(\R,+,0)$
to the monoidal structure $(\CC,\otimes,\I)$ associated with the
tensor product in $\CC$.  We choose to overload the names
$\tduplicate$ and $\tdrop$ here to accommodate for the generalization
from $\R$-comonads to $\R$-comonad algebras:
\[
\begin{array}{lcl}
  \tduplicate & : & \Hom {A_{r+s}} {A_r \otimes A_s} \\
  \tdrop      & : & \Hom {A_0} {\I} \\
\end{array}
\]
The laws of $\tduplicate$ and $\tdrop$ for comonadic objects
$\D[r]\,B$ immediately generalize to comonad algebras $A_r$:
\[
\xymatrix@C=12ex{
A_{0} \otimes A_r \ar[d]_{\tdrop \otimes \tid}
& A_r \ar[l]_{\tduplicate} \ar[r]^{\tduplicate} \ar@{=}[d]
& A_r \otimes A_{0} \ar[d]^{\tid \otimes \tdrop}
\\
\I \otimes A_r \ar[r]^{\Gl}
& A_r
& A_r \otimes \I \ar[l]_{\Gr}
}
\]
\[
\xymatrix@C=10ex{
A_q \otimes A_{r+s}  \ar[d]_{\tid \otimes \tduplicate}
& A_{q+r+s} \ar[l]_{\tduplicate} \ar[r]^{\tduplicate}
& A_{q+r} \otimes A_s \ar[d]^{\tduplicate \otimes \tid}
\\
A_q \otimes (A_r \otimes A_s) \ar[rr]^{\Ga}
& & (A_q \otimes A_r) \otimes A_s
}
\]
\[
\xymatrix@C=22ex{
A_{(q+r)s} \ar[r]^{\texpose} \ar[d]_{\tduplicate}
& \D[q+r]\,A_s \ar[d]^{\tduplicate}
\\
A_{qs} \otimes A_{rs} \ar[r]^{\texpose \otimes \texpose}
& \D[q]A_s \otimes \D[r]A_s
}
\]
\[
\xymatrix{
A_{qr} \otimes A_{qs} \ar[d]_{\expose \otimes \expose}
  & A_{q(r+s)} \ar[l]_{\tduplicate} \ar[r]^{\expose}
  & \D[q]\,A_{r+s} \ar[d]^{\fmap\tduplicate}
\\
\D[q]A_r \otimes \D[q]A_s \ar[rr]^{\tzip}
&&
\D[q](A_r \otimes A_s)
}
\]

@@

In coeffect-graded CBPV, we wish to interpret each type in the context
as a $\D$-coalgebra for a fixed graded comonad $\D$.  The
interpretation $\den \Gamma$ of the context $\Gamma$ should then be a
comonad coalgebra over the grading $\gamma$ such that judgements
$\qCompTy \gamma \Gamma t N$ can be interpreted as morphisms
$\dent t : \Hom {\Den\Gamma \gamma} {\denn N}$.  Since each variable
comes with its own resource qualifier, we cannot simply model the
context as a tensor product $\tupt {x:\dom\Gamma} \den{\Gamma(x)}$ of
$\D$-coalgebras since this product would indexed by a single resource
qualifier $r \in \R$ rather than a resource context
$\gamma : \dom\Gamma \to \R$.  The solution offered by
\citet{orchard:icfp14} are \emph{structured indexed comonads}.  We
shall generalize this to comonad coalgebras to the extend needed for
interpreting contexts.

>>>>

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
Call-by-push-value (CBPV) is a simply typed lambda calculus that polarizes
types into value and computation types and can thus express both
call-by-name and call-by-value evaluation in the presence of effects.
Semantically, effects are modeled by a monad, and computation types as
algebras over this monad.
%
Effect type systems usually express more information than the presence
of an effect; often effects are categorized by preordered monoid
where the monoid operation represents accumulation of effects and the
order expresses effect subsumption, in analogy to subtyping.
In this work, a \emph{graded} version of CBPV is presented where the typing
of computations likens effect typing.  Semantically, computation types
are then represented as graded monad algebras.
%
Observing that the value types of CBPV can be interpreted as comonad coalgebras, we further present a version of CBPV that has coeffects graded by a preordered semiring.  Value types and contexts are interpreted as graded comonad coalgebras, allowing resource-aware interpretations of CBPV.
%
Finally, we combine the two systems into a fully graded version of CBPV where both effects and coeffects are graded.  Thus turns out to be possible without specifying an interaction between effects and coeffects.
\end{abstract}

@@

%\newcommand{}{\ensuremath{}}
\newcommand{\bla}{\ensuremath{\mbox{$$}}}
\newcommand{\ie}{\emph{i.e.}}
\newcommand{\eg}{\emph{e.g.}}
\newcommand{\Eg}{\emph{E.g.}}
\newcommand{\loccit}{\emph{loc.\,cit.}\xspace}
\newcommand{\cf}{cf.\ }
\newcommand{\den}[2][]{\llbracket#2\rrbracket^{#1}}
\newcommand{\dent}[1]{\llparenthesis#1\rrparenthesis}
\newcommand{\To}{\ensuremath{\Rightarrow}}
%\newcommand{\todot}{\stackrel\cdot\to}
\newcommand{\todot}{\mathbin{\dot{\to}}}
\newcommand{\bN}{\ensuremath{\mathbb{N}}}
\newcommand{\dom}{\mathop{\mathrm{dom}}\nolimits}
\newcommand{\Pot}[1]{\mathcal{P}\,#1}
%\newcommand{\defiff}{:\iff}
\newcommand{\defiff}{\mathrel{{{:}{\Longleftrightarrow}}}}
\newcommand{\subst}[3]{#3[#1/#2]}
\newcommand{\qsubst}[4]{#4[#1/#2#3]}

@@

% Concrete effects
\newcommand{\tprint}{\mathsf{print}}
\newcommand{\print}[2]{\tprint\,#1.\,#2}
\newcommand{\toutput}{\mathsf{output}}
%\newcommand{\output}[2][]{\toutput_{#1}\,#2}
\newcommand{\tthrow}{\mathsf{throw}}
\newcommand{\throw}[1]{\tthrow\,#1}
\newcommand{\tcatch}{\mathsf{catch}}
\newcommand{\catch}[3]{#1\,\tcatch\,#2 \mapsto #3}
\newcommand{\traise}{\mathsf{raise}}
%\newcommand{\raise}[1]{\traise\,#1} %clash
\newcommand{\thandle}{\mathsf{handle}}
\newcommand{\handle}[4][]{\thandle_{#1}\,#2\,#3\,#4}

@@

Levy's call-by-push-value calculus (CBPV) \citeyearpar{levy:hosc06}
is a simply typed lambda
calculus with disjoint sums and eager and lazy products that allows
the modelling of ``lazy'' effects.  For instance, the CBPV-program
\[\print {\mathtt{"function"}} \lam x {\print {\mathtt{"argument"}} x}\]
of type $P \To \diamond P$, will, unlike the corresponding ML-program,
not print the word ``function'' and return a function; it will simply
wait for an input value of type $P$.  Only after an argument has been
supplied, the program will print the words ``function'' and
``argument'' and then return the argument.  The reason is that effects
in CBPV can only be observed at \emph{value types} like $P$, and
function types are not value types but computation types.  Effectful
actions (like printing) at computation types $N$ are ``pushed down''
the type structure until they reach a position at a monadic type
$\diamond P$ where they can be executed.  Semantically, this is
facilitated by interpreting computation types as monad algebras with
an action $\trun : \T\,N \to N$ that allows to formally run effects
transported by the monad $\T$ at type $N$, even though $N$ is not
directly a monadic type $\diamond P$.

CBPV polarizes types into value types $P$ and computation types $N$
where the former embed into the latter as $\diamond P$ by the formal
``monad''~$\diamond$ and the latter embed into the former as $\Box N$
by ``thunking''~$\Box$.  In Levy's \emph{behavioral semantics},
functor $\diamond$ is a left adjoint to functor $\Box$ and need not be
a monad.  Quite the opposite, when modelling store
$\diamond P = P \times S$ is a comonad and $\Box N = S \to N$ a monad.
Contrastingly, in the above explained \emph{algebra semantics},
$\diamond$ is indeed modeled by a monad $\T$, yet $\Box$ just by the
identity.  It is however possible to model $\Box$ by a monoidal
comonad $\D$, and positive types by comonad coalgebras with an action
$\texpose : P \to \D\,P$ that makes the services of the comonad
available at all value types, not just at $\Box N$.

Building on these observations, we investigate in this article how
CBPV %would look like if it would
can
keep track of not just the presence
of an effect or coeffect but also the kind of effect and coeffect.
Semantically, precise information about effects can be obtained by
using a family of monads $\T[e]$ graded over effect classifiers $e$
drawn from a preordered monoid \cite{katsumata:popl14}.  For CBPV, we
generalize the concept of monad algebras $N$ to \emph{graded} such
ones, $N_e$.  Syntactically, computation typing $\CompTy \Gamma t e N$
is extended by an effect classifier $e$
(as usual in type and effect systems \citep{nielsen:effect99})
such that a computation can be
interpreted by a morphism $\Gamma \to N_e$.

Coeffect type systems
\citep{brunel:esop14,orchard:icfp14,ghicaSmith:esop14} have been
developed to track resource consumption by attaching usage information
to each variable in the typing context.  Coeffect typing generalizes
linear typing \citep{girard:linear} to quantitative typing
\citep{sergeyVytiniotisPeytonJones:popl14,mcBride:wadler60,atkey:lics18}
and subsumes sensitivity analysis \citep{reedPierce:icfp10} and static
information control flow \citep{volpano:jcs96} aka security typing.
Coeffects can be modeled by a comonad $\D[r]$ graded over resource
qualifiers $r$ drawn from a preordered semiring.  For the sake of
coeffect-graded CBPV, we generalize graded comonads to graded comonad
coalgebras.  Value types $P$ and contexts $\Gamma$ are then
interpreted as such coalgebras.  Value typing
$\qValTy \gamma \Gamma v P$ and computation typing
$\qCompTy \gamma \Gamma t N$ are based on linear typing an come
equipped with a resource context~$\gamma$.  Semantically, values are
interpreted as morphisms $\Gamma_{r\gamma} \to P_r$ allowing the
``multiplication'' or \emph{scaling} of a value by $r$ % natural in $r$
and computations as morphisms $\Gamma_\gamma \to N$.

Finally, effect and coeffect graded CBPV can be combined into a fully
graded CBPV calculus, surprisingly without sorting out any interaction
between effects and coeffects, such as the distributive laws of
\citet{orchard:icfp16}.  We credit the smoothness of the integration
to the careful placement of monad and comonad in CBPV's type system,
and the lack of need for scaling % of not just values but also
of computations.

\paragraph*{Contributions}
\begin{enumerate}

\item We introduce graded monad algebras and an effect-graded version
  of CBPV in Section~\ref{sec:effect}, after recapitulating monads and
  their algebras and graded monad.

\item We further introduce graded comonad coalgebras and a
  coeffect-graded version of CBPV in Section~\ref{sec:coeff}.  We give
  its denotational and operational semantics and adapt Atkey and
  Wood's substitution theorem \citeyearpar{atkeyWood:types19} to this
  version of CBPV.

\item We present an effect- and coeffect-graded version of CBPV in Section~\ref{sec:full-cbpv}.

\end{enumerate}

\paragraph*{Preliminaries}  The reader should bring some elementary
knowledge of category theory, such as the interpretation of
simply-typed lambda calculus in cartesian-closed categories.  However,
we try to be gentle with categorical concepts such as monads and
recapitulate their definition where needed.  In many cases, it is
sufficient to think in terms of the category $\SET$ where objects are
sets and morphisms functions between sets, or in the functor category
$[\CC \to \SET]$ where objects are monotone families of sets
$(A_i)_{i:\CC}$ indexed by objects $i$ of $\CC$ such that
$\CC_i \to \CC_j$ in $\SET$ for $i \to j$ in $\CC$ and morphisms are
natural transformations $(f_i : A_i \to B_i)_{i:\CC}$.

\section{An effect-graded version of CBPV}
\label{sec:effect}

@@

@@

A \emph{monoidal} comonad implements a morphism $\tzip$ that combines
a tuples of comonadic values into one comonadic tuple:
\[
\begin{array}{lcl}
  \tzip & : & \tupt {i:I} {(\D\,B_i)} \to \D(\tupt I B) \\
  \tzip\,(s_i)_{i:I}\,n & = & (s_i\,n)_{i:I} \\
\end{array}
\]
Many other comonads are monoidal,
like the store comonad $(S \To \_) \times S$,
though not all, \eg, the context comonad $S{\times}\_$.
In this article, we consider only monoidal comonads.

@@

Because comonad coalgebras are closed under products for monoidal
coalgebras, contexts $\Gamma$ are interpreted as comonad coalgebras
$\den\Gamma$.  In the presence of a comonad interpretation of $\Box$,
the operations $\tthunk$ and $\tforce$ are no longer the identity, but
``generalized cobind'' and $\extract$:
\[
\begin{array}{lllcl}
 \rintro\Box
  & \ru{\oCompTy \Gamma t N
      }{\ValTy \Gamma {\thunk t} {\Box N}}
  & \dent{\thunk t} & = & \fmap[\D]{\dent t} \comp \expose[\den\Gamma]
\\[3ex]
 \relim\Box
  & \ru{\ValTy \Gamma v {\Box N}
      }{\oCompTy \Gamma {\force v} N}
  & \dent{\force v} & = & \textract \comp \dent v
\end{array}
\]
The map $\dent\tthunk : \Hom A B \to \Hom A {\D\,B}$ generalizes
$\tcobind : \Hom{\D\,C} B \to \Hom{\D\,C}{\D\,B}$ to $\D$-coalgebras
$A$ in the same way that the monadic
$\tbind : \Hom{A}{\T\,D} \to \Hom{\T\,A}{\T\,D}$ is generalized to
$\Hom{A}{B} \to \Hom{\T\,A}{B}$ for $\T$-algebras $B$ in CBPV.

@@

\emph{Graded comonads} have been utilized to give semantics to
context-dependent computation \citep{orchard:icfp14}.  For grading,
\loccit uses a \emph{resource algebra} in form of a preordered
semiring
$(\R,{+},0,{\cdot},1,{\leq})$.
% $(\R,\_{+}\_,0,\_{\cdot}\_,1,\leq)$.
%
The semantics is \emph{resource aware} and thus not constructed in a
cartesian-closed category but in a symmetric monoidal closed category
$(\CC,\otimes,\I,\lolli)$.
Introduction $\Lambda$ and elimination $\teval$ of
exponentials are there
$\Lambda : \Hom {C \otimes A} B \to \Hom C {A \lolli B}$
and $\teval : \Hom{(A \lolli B) \otimes A} B$.  The tensor product
$\otimes$ is a bifunctor and it is customary to overload $\otimes$ for
the functorial action
$f_1 \otimes f_2 : \Hom{A_1 \otimes A_2}{B_1 \otimes B_2}$ where
$f_i : \Hom{A_i}{B_i}$.  We shall also use $I$-ary products
$\tupt {i:I} {A_i}$ and their functorial action $\tupt {i:I} {f_i}$.
Further, the symmetric monoidal structure is usually witnessed by
natural
isomorphisms for left $\lambda_A : \Hom {\I \otimes A} A$ and right
unit $\rho_A : \Hom {A \otimes I} A$, associativity
$\Ga_{A,B,C} : \Hom{A \otimes (B \otimes C)} {(A \otimes B) \otimes
  C}$
and swap $\sigma_{A,B} : \Hom {A \otimes B} {B \otimes A}$.  We shall
summarize combinations of these isomorphisms under the name
$\isot : \Hom {\tupt I A} {\tupt J B}$ when the multisets
$\{A_i\}_{i:I}$ and $\{B_j\}_{j:J}$ coincide modulo addition and
deletion of units ($\I$).

As a running example for resource accounting,
we use the semiring $\R = \Pot\bN \setminus \emptyset$ of
\emph{multiplicities} ordered by $\supseteq$ with pointwise sum and
product, \eg, $r + s = \{ n + m \mid n \in r, m \in s \}$.
Subsemirings of $\R$ have been used for quantitative typing
\cite{sergeyVytiniotisPeytonJones:popl14,mcBride:wadler60,atkey:lics18}.
The order expresses precision of the quantities, \eg,
$\{1\} \geq \{0,1\} \geq \bN$ states that linear use of a variable is more informative than affine use than unrestricted use.

@@

Objects of $\CC$ are functors $A : [(\W,\wleq) \to \SET]$ and morphisms
$f : \Hom A B$ are natural transformations
$(f_w : A_w \to B_w)_{w:W}$, \ie,
$B_{w \wleq w'} \comp f_w = f_{w'} \comp A_{w \wleq w'}$.  The tensor
$A \otimes B$ is Day's convolution
$(A \otimes B)_w = \bigcup_{w_1 \oplus w_2 \wleq w} (A_{w_1} \times B_{w_2})$
% $(A \otimes B)_w = \bigcup_{w_1,w_2 \mid w \geq w_1 \oplus w_2} A_{w_1} \times A_{w_2}$
with unit $\I_w = \bigcup_{\emp \wleq w} 1$.  The unit $\I$ is
constructible at world $w$ from nothing (the unit set $1$) if $\emp \wleq w$,
\ie, if the world $w$ includes the choice of using no resources.  A
tensor $A \otimes B$ is constructible at world $w$ if $w$ includes the
choice to split the resources into $w_1$ and $w_2$ to construct $A$
and $B$, resp.  The exponential $A \lolli B$ is determined by currying
$C \otimes A \lolli B \cong C \lolli (A \lolli B)$ and given by
$(A \lolli B)_w = \bigcap_{w_1 \oplus w \wleq w_2} (A_{w_1} \to
B_{w_2})$.

The thus constructed symmetric monoidal closed category $\CC$ has a
$\R$-graded comonad
\[
\begin{array}{lcl}
  \D & : & [(\R,\leq) \to [\CC \to \CC]] \\
(\D[r] A)_w & = &
  \bigcup_{rw' \wleq w} A_{w'}
  % \bigcup_{w' \mid w \wleq rw'} A_{w'}
\end{array}
\]
implementing trivially the following natural transformations.
Herein, we use as second monoidal structure
on the functor category $[\CC \to CC]$ the
pointwise tensor product of functors
$(\dot\otimes_{i:I}F_i)\,A = \otimes_{i:I}(F_i\,A)$ in
$[\CC \to \CC]$, in particular $\dot\I\,A = \I$.
\[
\begin{array}{lcl@{~}c@{~}l}
  \textract   & : & \D[1]   & \todot & \Id \\
  \tdisplay & : & \D[rs]  & \todot & \D[r] \comp \D[s] \\
  \tdrop    & : & \D[0]   & \todot & \dot{\I} \\
  \tduplicate  & : & \D[r+s] & \todot & \D[r] \mathbin{\dot\otimes} \D[s] \\
  % \tdrop    & : & \Hom {\D[0]\,A} A \\
  % \tduplicate  & : & \Hom {\D[r+s]\,A} {\D[r]\,\A \otimes \D[s]\,A} \\
\end{array}
\]
For example, $\tduplicate$ takes
$a \in \bigcup_{rw' \oplus sw' \wleq w} A_{w'}$ to
$(a,a) \in \bigcup_{w_1 \oplus w_2 \wleq w} \bigcup_{rw_1' \leq w_1}
\bigcup_{sw_2' \leq w_2} A_{w'_1} \times A_{w'_2}$.
The intermediate worlds are $w_1' = w_2' = w'$ and $w_1 = rw'$ and $w_2 = sw'$.

@@

A $\D$-coalgebra is a functor $A : [(\R,\leq) \to \CC]$ with family of
morphisms $\expose[r,s] : A_{rs} \to \D[r]\,A_s$ natural in $r$ and
$s$.  For the $\D$-coalgebra $\D[\_]\,B$ the family $\expose$ is just
$\tdisplay$, and the suitably generalizable laws for $\tdisplay$ are
required to hold for $\texpose$:
\[
\xymatrix@C=10ex{
A_r  \ar[d]_{\texpose} \ar@{=}[dr]
\\
\D[1]\,A_r \ar[r]^{\textract}
  & A_r
}
\qquad
\xymatrix@C=12ex{
A_{qrs} \ar[d]_{\texpose} \ar[r]^{\texpose}
  & \D[qr]\,A_s \ar[d]^{\tdisplay}
\\
\D[q]\,A_{rs} \ar[r]^{\tfmap\,\texpose}
  & \D[q]\,(\D[r]\,A_s)
}
\]
Further, $A$ needs to map the additive monoidal structure $(\R,+,0)$
to the monoidal structure $(\CC,\otimes,\I)$ associated with the
tensor product in $\CC$.  We choose to overload the names
$\tduplicate$ and $\tdrop$ here to accommodate for the generalization
from $\R$-comonads to $\R$-comonad algebras:
\[
\begin{array}{lcl}
  \tduplicate & : & \Hom {A_{r+s}} {A_r \otimes A_s} \\
  \tdrop      & : & \Hom {A_0} {\I} \\
\end{array}
\]
The laws of $\tduplicate$ and $\tdrop$ for comonadic objects
$\D[r]\,B$ immediately generalize to comonad algebras $A_r$:
\[
\xymatrix@C=12ex{
A_{0} \otimes A_r \ar[d]_{\tdrop \otimes \tid}
& A_r \ar[l]_{\tduplicate} \ar[r]^{\tduplicate} \ar@{=}[d]
& A_r \otimes A_{0} \ar[d]^{\tid \otimes \tdrop}
\\
\I \otimes A_r \ar[r]^{\Gl}
& A_r
& A_r \otimes \I \ar[l]_{\Gr}
}
\]
\[
\xymatrix@C=10ex{
A_q \otimes A_{r+s}  \ar[d]_{\tid \otimes \tduplicate}
& A_{q+r+s} \ar[l]_{\tduplicate} \ar[r]^{\tduplicate}
& A_{q+r} \otimes A_s \ar[d]^{\tduplicate \otimes \tid}
\\
A_q \otimes (A_r \otimes A_s) \ar[rr]^{\Ga}
& & (A_q \otimes A_r) \otimes A_s
}
\]
\[
\xymatrix@C=22ex{
A_{(q+r)s} \ar[r]^{\texpose} \ar[d]_{\tduplicate}
& \D[q+r]\,A_s \ar[d]^{\tduplicate}
\\
A_{qs} \otimes A_{rs} \ar[r]^{\texpose \otimes \texpose}
& \D[q]A_s \otimes \D[r]A_s
}
\]
\[
\xymatrix{
A_{qr} \otimes A_{qs} \ar[d]_{\expose \otimes \expose}
  & A_{q(r+s)} \ar[l]_{\tduplicate} \ar[r]^{\expose}
  & \D[q]\,A_{r+s} \ar[d]^{\fmap\tduplicate}
\\
\D[q]A_r \otimes \D[q]A_s \ar[rr]^{\tzip}
&&
\D[q](A_r \otimes A_s)
}
\]

@@

In coeffect-graded CBPV, we wish to interpret each type in the context
as a $\D$-coalgebra for a fixed graded comonad $\D$.  The
interpretation $\den \Gamma$ of the context $\Gamma$ should then be a
comonad coalgebra over the grading $\gamma$ such that judgements
$\qCompTy \gamma \Gamma t N$ can be interpreted as morphisms
$\dent t : \Hom {\Den\Gamma \gamma} {\denn N}$.  Since each variable
comes with its own resource qualifier, we cannot simply model the
context as a tensor product $\tupt {x:\dom\Gamma} \den{\Gamma(x)}$ of
$\D$-coalgebras since this product would be indexed by a single resource
qualifier $r : \R$ rather than a resource context
$\gamma : \dom\Gamma \to \R$.  The solution offered by
\citet{orchard:icfp14} are \emph{structured indexed comonads}.  We
shall generalize this to comonad coalgebras to the extend needed for
interpreting contexts.
